```{r}
library(purrr)
library(rgbif)
library(dplyr)
library(stringr)
library(readr)
library(here)
library(progressr)
```


# Get taxa from checklists

In this chapter we select species checklists and retrieve the taxa they contain.

## Choose checklists {#choose-checklists}

The unified checklist is compiled from country checklists published to GBIF. Here we choose the checklists we want to include and rank them by trust (most trustworthy first). The ranking will help to choose between duplicate taxa in later steps.

1. Choose checklists:

```{r get_taxa-1, echo = TRUE}
checklist_keys <- c(
  "016c16c3-d907-4c88-97dd-97ad62c8130e" # andorra
  #"497fa25f-6a32-4a2a-8f42-d01d8a0b7c22" # cyprus
)
```

2. Add the [GBIF Backbone Taxonomy](https://doi.org/10.15468/39omei). **Note**: we won't use this as a source checklist, but we need its metadata in the Darwin Core mapping.

```{r get_taxa-2, echo = TRUE}
checklist_keys <- append(checklist_keys, "d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")
```

3. Get metadata for these checklists from GBIF and display the result:

```{r get_taxa-get_checklist_metadata_from_gbif, cache = TRUE}
checklists <-
  purrr::map(checklist_keys, function(x) rgbif::datasets(uuid = x)) %>%
  purrr::map_df(function(x) list(
      datasetKey = x$data$key,
      title = x$data$title,
      modified = x$data$modified,
      publisher = x$data$publishingOrganizationKey,
      doi = paste0("https://doi.org/", x$data$doi),
      citation = x$data$citation$text,
      license = x$data$license
    ))
checklists
```

4. Remove `accessed via GBIF.org on yyyy-mm-dd.` from citation (we want the static citation of the dataset).

```{r get_taxa-3}
checklists <-
  checklists %>% dplyr::mutate(citation = stringr::str_remove(citation, " accessed via GBIF.org on \\d+-\\d+-\\d+."))
```

5. Save to [CSV](https://github.com/onestop-project/unified-europe).

```{r get_taxa-4}
readr::write_csv(checklists, here::here("data", "raw", "checklists.csv"), na = "")
```

6. Remove the GBIF Backbone Taxonomy from further querying steps.

```{r get_taxa-5}
checklists <- head(checklists, -1)
```

## Get taxa

1. Get taxa from these checklists from GBIF. **Note**: here we get _checklist taxa_, not _GBIF backbone taxa_.

```{r get_taxa-get_checklist_taxa_from_gbif, cache = TRUE}
taxa <-
  purrr::map_df(checklists$datasetKey, function(x) {
    rgbif::name_usage(
      datasetKey = x,
      limit = 99999 # Should be more than total taxa for all checklists
    )$data
  })
```

2. Keep only source taxa, not denormed higher classification taxa (= taxa added by GBIF if `kingdom`, `phylum`, etc. was populated)

```{r get-taxa-6}
taxa <-
  taxa %>%
  filter(origin == "SOURCE")
```

3. Keep only taxa that are not considered synonyms by source checklist.

```{r get-taxa-7}
taxa <-
  taxa %>%
  filter(taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL"))
```

4. Select columns of interest, rename `key` to `taxonKey`.

```{r get_taxa-8}
taxa <-
  taxa %>%
  select(key, scientificName, taxonID, datasetKey, nameType, issues, nubKey) %>%
  rename(taxonKey = key)
```

5. Fix `scientificName` spelling issues (i.e. double quotes).

```{r get_taxa-9}
taxa <-
  taxa %>%
  mutate(
    scientificName = str_replace_all(scientificName, "\"", "'")
  )
```

6. Preview checklist taxa:

```{r get_taxa-10}
taxa %>% head()
```

## Filter on distributions {#filter-on-distribution}

Checklists may contain more than alien species. We therefore need to filter on the associated [distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml) information.

**Note**: we filter on distribution information of _checklist taxa_, not _GBIF backbone taxa_. That is because backbone taxa contain distribution information from country checklists _and_ other checklists, which we don't want to consider here.

E.g. compare:

* Distributions for _checklist_ taxon [Eriocheir sinensis](https://api.gbif.org/v1/species/140563012/distributions) (`140563012`)
* Distributions for _backbone_ taxon [Eriocheir sinensis](https://api.gbif.org/v1/species/2225776/distributions) (`2225776`)

**Note**: if a checklist has related information for a taxon, but not a valid distribution, that related information will NOT be included in the unified checklist. 

1. Get distributions for our taxa from GBIF.

```{r get_taxa-get_distributions_from_gbif, cache = TRUE}
# Extract taxonKeys as a vector
taxon_keys <-
  taxa %>%
  pull(taxonKey)

# Get distributions
progressr::with_progress({
  progress_bar <- progressr::progressor(steps = length(taxon_keys))
  distributions <-
    furrr::future_map_dfr(
      taxon_keys,
      function(x) {
        progress_bar()
        rgbif::name_usage(
          key = x,
          data = "distribution"
        )$data
      },
      .options = furrr::furrr_options(seed = TRUE)
    )
})
```

2. Filter distributions on present, alien species:

```{r get_taxa-11, echo = TRUE}
distribution <-
  distributions %>%
  filter(
    establishmentMeans %in% c("INTRODUCED", "NATURALISED", "INVASIVE", "ASSISTED COLONISATION")
    #!status %in% c("ABSENT", "EXCLUDED", "DOUBTFUL") # Inverse filter!
  )
```

3. Save distributions to [CSV](https://github.com/onestop-project/unified-europe).

```{r get_taxa-12}
write_csv(distributions, here("data", "raw", "distributions.csv"), na = "")
```

4. Based on the filtered distributions, assign a `validDistribution` (`TRUE`/`FALSE`) column to the `taxa`.

```{r get_taxa-13}
taxa <-
  taxa %>%
  mutate(validDistribution = if_else(
    taxonKey %in% unique(distributions$taxonKey),
    TRUE,
    FALSE
  ))
```

5. Preview some taxa with not a single valid distribution (`taxonKey` can be used to verify manually on GBIF):

```{r get_taxa-14}
taxa %>%
  filter(!validDistribution) %>%
  head()
```

6. Rename `nubKey` to `bb_key` and move to the end.

```{r get_taxa-15}
taxa <-
  taxa %>%
  rename(bb_key = nubKey) %>%
  select(-bb_key, everything())
```

## Get GBIF backbone taxonomy information

A taxon can occur on more than one checklist. To identify these duplicates, we cannot rely on the _scientific name_ as there might be spelling variations (e.g. with or without authorship) and it does not account for synonyms that should be lumped with the accepted taxon. To have **unifying taxon identifiers across taxa**, we rely on the [GBIF backbone taxonomy](https://doi.org/10.15468/39omei), to which every checklist taxon (re)published to GBIF is automatically matched. If a match in the backbone is found, the checklist taxon will have a `nubKey`.

[To be continued]