---
title: "Download taxa and distributions"
author: "Sanne Govaert"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

## Install elodea

```{r}
pak::pak("onestop-project/elodea")
```

## Load libraries

```{r}
library(elodea)
library(yaml)
library(here)
library(purrr)
```

## Read yaml

In the `checklists.yml` file, we have a list of national checklists with their corresponding `datasetKey` and country name. We read this file into R using the `yaml::read_yaml()` function. This returns a list of national checklists that we can then use to download the taxa and distributions for each country.

```{r}
checklists <- yaml::read_yaml(here::here("src", "checklists.yml"))
```

## Filter checklists if needed

Choose here which checklists you want to download. Be aware that downloading all checklists takes a long time.

```{r}
to_keep <- c(
  "Albania",
  "Andorra",
  "Armenia",
  "Austria",
  "Azerbaijan",
  "Belarus",
  "Belgium",
  "Bosnia and Herzegovina",
  "Bulgaria",
  "Croatia",
  "Cyprus",
  "Cyprus - Sovereign Base Areas",
  "Cyprus - oneSTOP list",
  "Czechia",
  "Denmark",
  "Estonia",
  "Finland",
  "France",
  "Germany",
  "Greece",
  "Hungary",
  "Iceland",
  "Ireland",
  "Italy",
  "Latvia",
  "Liechtenstein",
  "Lithuania",
  "Luxembourg",
  "Malta",
  "Moldova",
  "Monaco",
  "Montenegro",
  "Netherlands",
  "North Macedonia",
  "Norway",
  "Poland",
  "Portugal",
  "Portugal - Azores",
  "Portugal - Madeira",
  "Romania",
  "San Marino",
  "Serbia",
  "Slovakia",
  "Slovenia",
  "Spain",
  "Sweden",
  "Switzerland",
  "Turkiye",
  "Ukraine",
  "United Kingdom"
)

checklists <- purrr::keep(checklists, ~ .x$country %in% to_keep) 
```

# Download taxa and distributions

This function downloads the taxa and distributions for a given `datasetKey` and saves them as CSV files in the specified directory (see next code block). The directory is named after the country, and the files are named `taxa.csv` and `distributions.csv`. The function also prints informative messages about the progress of the download and file writing process.

```{r}
save_taxa_and_distributions <- function(datasetKey, directory) {
  country <- basename(directory)
  cli::cli_h1(country)
  cli::cli_h2("Downloading files")
  cli::cli_ul("Downloading taxa")
  taxa <- elodea::get_taxa(datasetKey)
  cli::cli_ul("Downloading distributions")
  distributions <- elodea::get_distributions(datasetKey, taxa)
  taxa_path <- file.path(directory, "taxa.csv")
  distributions_path <- file.path(directory, "distributions.csv")
  cli::cli_h2("Writing files")
  cli::cli_ul("{.file {taxa_path}}")
  cli::cli_ul("{.file {distributions_path}}")
  if (!dir.exists(directory)) {
    dir.create(directory, recursive = TRUE)
  }
  readr::write_csv(taxa, taxa_path, na = "")
  readr::write_csv(distributions, distributions_path, na = "")
}
```

As a next step, we loop through the list of countries (`checklists`) and call the `save_taxa_and_distributions()` function for each country, passing the corresponding `datasetKey` and directory path.
The directory is `data/interim/{country}` where `{country}` is the name of the country as specified in the `checklists.yml` file.

```{r}
purrr::walk(
  checklists,
  ~ save_taxa_and_distributions(
    datasetKey = purrr::pluck(.x, "uuid"),
    directory = here::here("data", "interim", purrr::pluck(.x, "country"))
  )
)
```