---
title: "Filter local data"
author: "Sanne Govaert"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

This script 
- reads the downloaded taxa and distributions files for each country from `data/interim`
- clean the data by filtering
- writes the filtered data and notes to `data/processed`

See [elodea::filter_data()](https://onestop-project.github.io/elodea/reference/filter_data.html) for details on filtering and notes.

## Load libraries

```{r}
library(elodea)
library(here)
library(readr)
```

## Set directory

Set the directory to `data/interim`, which needs to contain the downloaded taxa and distributions files for each country.

```{r}
parent_dir <- here::here("data", "interim")
```

### Select specific countries if needed

```{r}
# List all folders (countries) in the parent directory
country_folders <- list.dirs(parent_dir, recursive = FALSE)

# Select specific countries if needed
countries <- c(
  "Albania",
  "Andorra",
  "Armenia",
  "Austria",
  "Azerbaijan",
  "Belarus",
  "Belgium",
  "Bosnia and Herzegovina",
  "Bulgaria",
  "Croatia",
  "Cyprus",
  "Cyprus - Sovereign Base Areas",
  "Cyprus - oneSTOP list",
  "Czechia",
  "Denmark",
  "Estonia",
  "Finland",
  "France",
  "Germany",
  "Greece",
  "Hungary",
  "Iceland",
  "Ireland",
  "Italy",
  "Latvia",
  "Liechtenstein",
  "Lithuania",
  "Luxembourg",
  "Malta",
  "Moldova",
  "Monaco",
  "Montenegro",
  "Netherlands",
  "North Macedonia",
  "Norway",
  "Poland",
  "Portugal",
  "Portugal - Azores",
  "Portugal - Madeira",
  "Romania",
  "Slovakia",
  "Slovenia",
  "Spain",
  "Sweden",
  "Switzerland",
  "Turkiye",
  "Ukraine",
  "United Kingdom"
)

selected_paths <- country_folders[basename(country_folders) %in% countries]
selected_paths
```

## Filter data

### Define a function to process one folder

This function takes a folder path and a directory path as input. It reads the `taxa.csv` and `distributions.csv` files from the folder, filters the data to keep only the introduced species using the `elodea::filter_data()` function, and writes the filtered data and notes to the specified directory under a subfolder named after the country.

```{r}
# Define a function to process one folder
process_folder <- function(folder, directory) {
  country <- basename(folder)
  
  taxa_file <- file.path(folder, "taxa.csv")
  distributions_file <- file.path(folder, "distributions.csv")
  
  taxa <- readr::read_csv(taxa_file, show_col_types = FALSE)
  distributions <- readr::read_csv(distributions_file, show_col_types = FALSE)
  
  result <- elodea::filter_data(taxa, distributions, establishment_means = "introduced")
  
  dir <- file.path(directory, country)
  taxa_path <- file.path(dir, "taxa.csv")
  distributions_path <- file.path(dir, "distributions.csv")
  notes_path <- file.path(dir, "notes.csv")
  if (!dir.exists(dir)) {
    dir.create(dir, recursive = TRUE)
  }
  cli::cli_h2("Writing files for {country}")
  readr::write_csv(result$taxa, taxa_path, na = "")
  cli::cli_ul("{.file {taxa_path}}")
  readr::write_csv(result$distributions, distributions_path, na = "")
  cli::cli_ul("{.file {distributions_path}}")
  readr::write_csv(result$notes, notes_path, na = "")
  cli::cli_ul("{.file {notes_path}}")
}
```

### Process all folders

Loop through the selected folders and call the `process_folder()` function for each folder, passing the corresponding directory path. The processed files will be written to `data/processed/{country}` where `{country}` is the name of the country.

```{r}
# Use purrr::map to process all folders
purrr::map(
  selected_paths, 
  ~ process_folder(.x, directory = here::here("data", "processed"))
  )
```


